String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-02 2023-1-19/sections/abstract.tex
experiments
remove under approximation

Regular expressions (regex for short) and string-length function are widely used in string-manipulating programs. 
%Solving string constraints involving regular membership and string-length function has been widely investigated. 
Counting is a frequently used feature of regexes that count the number of matchings of sub-patterns. The state-of-the-art string solvers are incapable of solving string constraints with regex-counting and string-length efficiently, especially when the counting bounds are large. In this work, we propose an automata-theoretic approach for solving such class of string constraints efficiently. 
%
The main idea is to model the counting operators symbolically by registers in automata, instead of unfolding them explicitly, thus alleviating the state explosion problem resulted from the unfolding.  
%
Moreover, we model the string lengths by registers as well. 
%Since Parikh images of automata can be defined by linear integer arithmetic formulas, 
As a result, the satisfiability of string constraints with regex-counting and string-length is reduced to the satisfiability of linear integer arithmetic, which can then be solved by off-the-shelf SMT solvers. 
%
%we extract linear integer arithmetic formulas from automata, which encode the constraints on the values of registers, and utilize the off-the-shelf SMT solvers to solve the linear integer arithmetic constraints. 
%
To improve the performance further, on the one hand, we propose techniques to reduce automata sizes, 
%thus reducing the sizes of linear arithmetic formulas computed from automata as well, 
and on the other hand, we utilize under approximations to search for solutions.   
% we propose state reduction techniques and under-approximation to improve the performance further. 
We implement the algorithms and validate our approach on 48,834 benchmark instances. The experimental results show that 
our approach can solve more instances than the other tools, with comparable or better performance.
% performance of this approach compared against the state-of-the-art string solvers.

>>>>

Regular expressions (regex for short) and string-length function are widely used in string-manipulating programs. 
%Solving string constraints involving regular membership and string-length function has been widely investigated. 
Counting is a frequently used feature of regexes that count the number of matchings of sub-patterns. The state-of-the-art string solvers are incapable of solving string constraints with regex-counting and string-length efficiently, especially when the counting bounds are large. In this work, we propose an automata-theoretic approach for solving such class of string constraints efficiently. 
%
The main idea is to model the counting operators symbolically by registers in automata, instead of unfolding them explicitly, thus alleviating the state explosion problem resulted from the unfolding.  
%
Moreover, we model the string lengths by registers as well. 
%Since Parikh images of automata can be defined by linear integer arithmetic formulas, 
As a result, the satisfiability of string constraints with regex-counting and string-length is reduced to the satisfiability of linear integer arithmetic, which can then be solved by off-the-shelf SMT solvers. 
%
%we extract linear integer arithmetic formulas from automata, which encode the constraints on the values of registers, and utilize the off-the-shelf SMT solvers to solve the linear integer arithmetic constraints. 
%
To improve the performance further, we propose techniques to reduce automata sizes.
% we propose state reduction techniques and under-approximation to improve the performance further. 
We implement the algorithms and validate our approach on 48,834 benchmark instances. The experimental results show that 
our approach can solve more instances than the other solvers, at a comparable or faster speed.
% performance of this approach compared against the state-of-the-art string solvers.

### 2023-08-02 2023-1-19/sections/algorithms.tex
remove under approximation

Finally, although we propose techniques to reduce the sizes of $\cU_{\cB_i}$'s, it is still relatively expensive to compute LIA formulas from $\cC_i$'s and solve their satisfiability, especially for those satisfiable problem instances, namely, those instances where CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty w.r.t. $\varphi$.   
Therefore, in the sequel, we propose a heuristic to search for solutions by utilizing under approximations, to be executed before Step 3. Intuitively, we search for solutions by bounding the lengths of the strings that are accepted by $\cC_i$'s. The heuristic uses a constant bound $B > 0$ and searches for solutions incrementally, starting from strings of length $0$. Since the heuristic is executed after Step 2 and before Step 3, we call it Step 2.5.

For $i \in [n]$, suppose $R_{\cB_i} = (r_{i,1},\cdots, r_{i, k_i})$ and $\cC_i = (Q'_i, \costset_{\cC_i}, \delta'_i, q_{i,0}, F'_i)$.

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let 
%$\confset_{\cC_i}:=\emptyset$, 
$\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ 
and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure until $len = B+1$.
\begin{enumerate}
\item Let $\xi: = \bigwedge \limits_{i \in [n]} \alpha_i \wedge \varphi$.
%
\item For each $i \in [n]$, do the following.
\begin{enumerate}
\item Let 
%$\confset_{\cC_i}:= \confset_{\cC_i} \cup \confset'_{\cC_i}$, 
$old_i:=\confset_{\cC_i}$, $\confset_{\cC_i}:=\emptyset$.
%
\item For each $(q, \myvec{v}) \in old_i$,
\begin{enumerate}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
%\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset_{\cC_i}:=\confset_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{enumerate}
%
%\item 
%
\item Let $\xi := \xi \wedge \psi_{\cC_i}$. 
\end{enumerate}
%
\item If $\xi$ is satisfiable, then report $\ltrue$ and the whole algorithm terminates. 
%
\item Let $len: = len+1$. \qed
\end{enumerate}
Intuitively, for each $i \in [n]$, $\psi_{\cC_i}$ is a global variable whose value is never reset, and it stores the set of ``final'' valuations of registers of $\cC_i$ (that is, the valuations obtained when reaching final states), and $\confset_{\cC_i}$ stores the set of newly discovered (not necessarily final) ``configurations'' of $\cC_i$.
%, and $\confset_{\cC_i}$ stores those that have been discovered before. 
Note that the solution searching algorithm in Step 2.5 is incremental because when the value of $len$ is incremented, the searching continues from $\confset_{\cC_i}$'s, not from scratch.
%, whose values are not reset.
%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

>>>>

% Finally, although we propose techniques to reduce the sizes of $\cU_{\cB_i}$'s, it is still relatively expensive to compute LIA formulas from $\cC_i$'s and solve their satisfiability, especially for those satisfiable problem instances, namely, those instances where CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty w.r.t. $\varphi$.   
% Therefore, in the sequel, we propose a heuristic to search for solutions by utilizing under approximations, to be executed before Step 3. Intuitively, we search for solutions by bounding the lengths of the strings that are accepted by $\cC_i$'s. The heuristic uses a constant bound $B > 0$ and searches for solutions incrementally, starting from strings of length $0$. Since the heuristic is executed after Step 2 and before Step 3, we call it Step 2.5.

% For $i \in [n]$, suppose $R_{\cB_i} = (r_{i,1},\cdots, r_{i, k_i})$ and $\cC_i = (Q'_i, \costset_{\cC_i}, \delta'_i, q_{i,0}, F'_i)$.

% \paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let 
% %$\confset_{\cC_i}:=\emptyset$, 
% $\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ 
% and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure until $len = B+1$.
% \begin{enumerate}
% \item Let $\xi: = \bigwedge \limits_{i \in [n]} \alpha_i \wedge \varphi$.
% %
% \item For each $i \in [n]$, do the following.
% \begin{enumerate}
% \item Let 
% %$\confset_{\cC_i}:= \confset_{\cC_i} \cup \confset'_{\cC_i}$, 
% $old_i:=\confset_{\cC_i}$, $\confset_{\cC_i}:=\emptyset$.
% %
% \item For each $(q, \myvec{v}) \in old_i$,
% \begin{enumerate}
% \item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
% %
% %\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
% %
% \item for each $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset_{\cC_i}:=\confset_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
% \end{enumerate}
% %
% %\item 
% %
% \item Let $\xi := \xi \wedge \psi_{\cC_i}$. 
% \end{enumerate}
% %
% \item If $\xi$ is satisfiable, then report $\ltrue$ and the whole algorithm terminates. 
% %
% \item Let $len: = len+1$. \qed
% \end{enumerate}
% Intuitively, for each $i \in [n]$, $\psi_{\cC_i}$ is a global variable whose value is never reset, and it stores the set of ``final'' valuations of registers of $\cC_i$ (that is, the valuations obtained when reaching final states), and $\confset_{\cC_i}$ stores the set of newly discovered (not necessarily final) ``configurations'' of $\cC_i$.
% %, and $\confset_{\cC_i}$ stores those that have been discovered before. 
% Note that the solution searching algorithm in Step 2.5 is incremental because when the value of $len$ is incremented, the searching continues from $\confset_{\cC_i}$'s, not from scratch.
%, whose values are not reset.
%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

### 2023-08-02 2023-1-19/sections/introduction.tex
remove under approximation

According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers for LIA. 
%
Because the sizes of LIA formulas depend much on the sizes of CEFA, it is beneficial to reduce the sizes of CEFA, in order to achieve a good performance. As a result, we propose various techniques to reduce the state space and the number of transitions of CEFA. Furthermore, since the computation of LIA formulas from CEFA is relatively expensive, we also propose a method of searching for strings that are accepted by CEFA, based on under approximations, before constructing the LIA formulas.

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 48,834 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,834 instances, while Z3str3RE, the state-of-the-art best solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA w.r.t. LIA formulas was solved by utilizing the nuXmv model checker~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh-image computation, automata-size reduction, and under approximation. The experimental results show that our approach solves 7,092 more instances than the solver in~\cite{atva2020} while spending less time. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details, where OSTRICH denotes the solver in \cite{atva2020}.)

>>>>

According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers for LIA. 
%
Because the sizes of LIA formulas depend much on the sizes of CEFA, it is beneficial to reduce the sizes of CEFA, in order to achieve a good performance. As a result, we propose various techniques to reduce the state space and the number of transitions of CEFA.

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 48,834 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,834 instances, while Z3str3RE, the state-of-the-art best solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA w.r.t. LIA formulas was solved by utilizing the nuXmv model checker~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh-image computation and automata-size reduction. The experimental results show that our approach solves 7,092 more instances than the solver in~\cite{atva2020} while spending less time. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details, where OSTRICH denotes the solver in \cite{atva2020}.)

### 2023-08-02 2023-1-19/sections/overview.tex
remove under approximation

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from NFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially, we set $\ell$, an integer variable to denote the lengths of strings, to $1$. If there does not exist an $\ell$-bounded satisfiable assignment, then we let $\ell:=\ell+1$ and continue the search, until finding a satisfiable assignment or reaching the bound $B$. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $3$ is sufficient to find a solution for the string constraint in~(\ref{eqn-running}), which is witnessed by the following accepting run of $\cC$: $q^{\prime\prime}_0 \xrightarrow{(1,0,0,1)} q^{\prime\prime}_0 \xrightarrow{(0,1,0,1)} q^{\prime\prime}_1 \xrightarrow{(0,0,0,1)} q^{\prime\prime}_2$. Note that the accepting run produces $(1,1,0,3)$, a vector of register values,  which satisfies $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. (See Section~\ref{subsec:cefadec} for more details.)

>>>>

% At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from NFA and solving them by SMT solvers are still expensive.  
% Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially, we set $\ell$, an integer variable to denote the lengths of strings, to $1$. If there does not exist an $\ell$-bounded satisfiable assignment, then we let $\ell:=\ell+1$ and continue the search, until finding a satisfiable assignment or reaching the bound $B$. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $3$ is sufficient to find a solution for the string constraint in~(\ref{eqn-running}), which is witnessed by the following accepting run of $\cC$: $q^{\prime\prime}_0 \xrightarrow{(1,0,0,1)} q^{\prime\prime}_0 \xrightarrow{(0,1,0,1)} q^{\prime\prime}_1 \xrightarrow{(0,0,0,1)} q^{\prime\prime}_2$. Note that the accepting run produces $(1,1,0,3)$, a vector of register values,  which satisfies $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. (See Section~\ref{subsec:cefadec} for more details.)
